Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-04-25 paper/main.tex
discussion, citation

\section{Introduction}
For many tasks, it is necessary to have DNA regions labeled according to their function. This process is called gene annotation and it can be performed in different levels of precision, from simple coding -- non-coding classification to detailed structural labeling, such as enhancer regions or long non-coding RNA sequences in eukaryotes. In this work on eucaryotic genomes, we differentiate between Intergenic regions, Untranslated regions (UTRS, parts at the beginning and end of gene, that are transcribed but not translated), Coding sequences (CDS) and Introns. Because of the sheer size of genomes alone, gene annotation is a very labour intensive process and genomic database providers like NCBI \citep{thibaud2013eukaryotic} and Ensembl \citep{aken2016ensembl} offer gene annotation pipelines. These pipelines integrate experimental data (from e.g. RNA-seq or proteogenomics) with homolog sequences in the database and ab-initio gene predictions. The latter is an attractive approach, because it is cheap and fast. State of the art performance ab-initio gene prediction is achieved by higher order hidden markov models (HMMs), such as Genscan \citep{burge1997prediction} or Augustus \citep{stanke2003gene} but their accuracy leaves room for improvement. By choosing possible states and transitions, designers of HMMs assume structure in the sequence that may limit its predictive power. Indeed, in practice HMMs have problems with generalizing across species and it seems actual learning of underlying patterns is limited. In the last decade, deep neural networks have been applied with tremendous success in many areas of statistical modelling. For sequence data, such as DNA, recurrent neural networks (RNN) in the form of long-short term memory (LSTM) are an established architecture and here, we employ them to improve on HMMs for ab-initio genome annotation.

@@

AUGUSTUS does outperform us rather consistently on the smallest genomes, but struggles much more the larger the genomes get. The difference in prediction quality is especially strong for mammals, which tend to have large genomes with very long genes, and the largest plants.

@@

Different set of training genomes could lead to a shift in strengths and weaknesses of the model while a larger and more spread out set of high quality genomes could result in a wider range of genomes with decent predictions.

>>>>

\section{Introduction}
For many tasks, it is necessary to have DNA regions labeled according to their function. This process is called gene annotation and it can be performed in different levels of precision, from simple coding -- non-coding classification to detailed structural labeling, such as enhancer regions or long non-coding RNA sequences in eukaryotes. In this work on eucaryotic genomes, we differentiate between Intergenic regions, Untranslated regions (UTRS, parts at the beginning and end of gene, that are transcribed but not translated), Coding sequences (CDS) and Introns. Because of the sheer size of genomes alone, gene annotation is a very labour intensive process and genomic database providers like NCBI \citep{thibaud2013eukaryotic} and Ensembl \citep{aken2016ensembl} offer gene annotation pipelines. These pipelines integrate experimental data (from e.g. RNA-seq or proteogenomics) with homolog sequences in the database and ab-initio gene predictions. The latter is an attractive approach, because it is cheap and fast. State of the art performance ab-initio gene prediction is achieved by higher order hidden markov models (HMMs), such as Genscan \citep{burge1997prediction} or Augustus \citep{stanke2003gene} but their accuracy leaves room for improvement. By choosing possible states and transitions, designers of HMMs assume structure in the sequence that may limit its predictive power. Indeed, in practice HMMs have problems with generalizing across species and it seems actual learning of underlying patterns is limited. In the last decade, deep neural networks have been applied with tremendous success in many areas of statistical modelling, including biology \citep{ching2018opportunities}. For sequence data, such as DNA, recurrent neural networks (RNN) in the form of long-short term memory (LSTM) are an established architecture and here, we employ them to improve on HMMs for ab-initio genome annotation.

@@

AUGUSTUS does outperform us rather consistently on the smallest genomes, but falls off much more with growing genome length. The difference in prediction quality is especially strong for mammals, which tend to have large genomes with very long genes, and the largest plants.

@@

We found our models to be highly sensitive to the training genomes we chose. A different set could lead to a significant shift in strengths and weaknesses of the model and a larger and more spread out set of high quality genomes could also result in a wider range of genomes with decent predictions. For this, more computational resources would be required.

An avenue for future research could be the addition of RNASeq data as additional input. This would bring Helixer on even footing with current tools and could lead to real world applicable performance improvements if the results of this work are any indication, as deep learning has been shown to excel in a multimodal settings \citep{ching2018opportunities}.
